---
title: "02_DecontamCOIB"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

The next steps of the workflow tend to throw more errors, rely on packages that may be unstable, and/or just require a bit more attention. Therefore these steps are placed in a .rmd for ease of troubleshooting and documenting notes. To run a chunk of code, just press the green play button at the top right side of each chunk.

This .rmd will only finish processing for the COIB primer dataset. See 02_Decontam12S.Rmd to finish sample processing for the 12S primer dataset.


```{r load, echo = FALSE}
# Load config file
source("R/01_config.R")

# Load the seqtabNC file - make sure the output versions match
seqtabB_NC <- readRDS("Data/Outputs/2025-06-17/COIB/InterimObjects/MergedSeqTab_NoChim.rds")
```


# Step 1: Assign Taxonomy to the DADA2 Output
This step may take some time
```{r assign_taxonomy, echo = FALSE}
all.taxa <- assignTaxonomy(seqtabB_NC, 'Data/Input/COI_REFDB.fasta', tryRC = TRUE, verbose = TRUE)

beep(sound = "fanfare")
```

# Step 2: Transfer to Phyloseq

To build a phyloseq object, key checks must be met. In the all.taxa object, the row names should be the sequences, and the meta and seqtabNC objects should match sample names by both character strings and order of samples. For whatever reason, assignTaxonomy has a habit of returning objects that are slightly different in layout each time I run it, but as long as those two key checks are met, it should work.

First we check all.taxa and the column name assignments to ensure the correct taxonomic ranks are labelled.
```{r taxacheck, echo = FALSE}
head(all.taxa)

# To correct the column names:
colnames(all.taxa) <- c("Phylum", "Class", "Order", "Family", "Genus", "Species")

# Save all.taxa as a .csv once itʻs in order
write.csv(all.taxa, file.path(path_outB, "COIBAssignedTaxa.csv"))

# If the row.names in all.taxa are not the sequences, remove the # and run the following line:
# all.taxa <- data.frame(all.taxa, row.names = 1)
```

Now check seqtabNC (primarily by organizing it) - seqtabNC should have the samples as rownames and sequences as colnames
```{r seqtabcheck, echo = FALSE}
head(seqtabB_NC) 
seqtabB_NC <- seqtabB_NC[order(row.names(seqtabB_NC)), ]

row.names(seqtabB_NC) <- gsub("_BF1.fastq.gz", "", row.names(seqtabB_NC)) # remove ending to names
```

And check the metadata file by removing the rows of samples that did not make it to seqtabNC, then set the sample names as rownames
```{r metacheck, echo = FALSE}
head(meta) # take a look at how meta read in

# Check which samples are in meta but not in seqtabM_NC - record this output
setdiff(meta$sample, row.names(seqtabB_NC))
```


```{r setdiff_and_clean, echo = FALSE}
# Remove the samples that were listed from setdiff() above
meta <- subset(meta, sample != "AMC22_AB01") 
meta <- subset(meta, sample != "AMC23_AB04") 
meta <- subset(meta, sample != "AMC23_MB09") 

# Set the sample column as row.names
meta <- data.frame(meta, row.names = 1)
# Order the meta file by row.names
meta <- meta[order(row.names(meta)), ]
```


With the rownames matching, we can now make the phyloseq object
```{r ps_object, echo = FALSE}
# Create phyloseq object with all samples
phy_ini <- phyloseq(otu_table(seqtabB_NC, taxa_are_rows=FALSE), sample_data(meta), tax_table(as.matrix(all.taxa)))

# SAVE 
saveRDS(phy_ini, "Data/Outputs/2025-06-17/COIB/InterimObjects/InitialPhyloseq.rds")
```

### Optional: Explore patterns of read abundance before proceeding to decontamination and taxa-filtering
```{r initial_exploratory, echo = FALSE}
# This chunk will make a stacked bar plot of detected reads per sample by Phylum assignment

# Set function to make a proportional graph
phy_ini.func <- transform_sample_counts(phy_ini, function(x) x / sum(x) )

# Plot based on Phylum
plot_bar(phy_ini.func, fill="Phylum") + ggtitle("Proportion by Phylum of Identified Reads - COIB")

# Save plot
ggsave("PreDecontam_IDProportions.png", path = path_figB, plot_bar(phy_ini.func, fill="Phylum") + ggtitle("Proportion by Phylum of Identified Reads"), width = 7, height = 4, units = "in") 
```

```{r library_size, echo = FALSE}
# This chunk will make a scatterplot of the number of reads by sample

# Put sample data into a ggplot-friendly data frame
df_ini <- as.data.frame(sample_data(phy_ini))

# Attach the read totals by sample to the data frame
df_ini$LibrarySize <- sample_sums(df_ini)

# Order the data frame by library size
df_ini <- expdf[order(df_ini$LibrarySize),]

# Create an index to number each row
df_ini$Index <- seq(nrow(df_ini))

# Make the plot
ggplot(data=df_ini, aes(x=Index, y=LibrarySize, color=Type)) + geom_point()

# Save the plot
ggsave("DecontamLibrarySize_Type.png", path = path_figB, ggplot(data=df_ini, aes(x=Index, y=LibrarySize, color=Type)) + geom_point() + ggtitle("COIB Identified Read Numbers per Sample, by Sample Type"), width = 6, height = 4, units = "in") 
```

# Step 3: Decontam

The standard decontam package uses either frequency or prevalence to identify and remove contaminants, but relies on blanks being true negative controls to use presence/absence of reads between true and negative samples to identify contaminants. Since these blank samples do have a significant number of reads, I use the Ishaq decontam method instead - isolating reads that were identified in the controls and removing them from all samples, based on which samples were extracted with which blank.

```{r decontam_isolate, echo = FALSE}
# subset samples by their DNA extraction batch
batch1 = subset_samples(phy_ini, Ebatch == "1")  
batch2 = subset_samples(phy_ini, Ebatch == "2")
batch3 = subset_samples(phy_ini, Ebatch == "3")  
batch4 = subset_samples(phy_ini, Ebatch == "4")  
batch5 = subset_samples(phy_ini, Ebatch == "5")  
batch6 = subset_samples(phy_ini, Ebatch == "6")  
batch7 = subset_samples(phy_ini, Ebatch == "7")  
batch8 = subset_samples(phy_ini, Ebatch == "8")  
batch9 = subset_samples(phy_ini, Ebatch == "9")  
batch10 = subset_samples(phy_ini, Ebatch == "10")  


# subset controls in each batch and prune to only those taxa
batch1_sub = subset_samples(batch1, Type == "NegCon")
batch2_sub = subset_samples(batch2, Type == "NegCon")
batch3_sub = subset_samples(batch3, Type == "NegCon")
batch4_sub = subset_samples(batch4, Type == "NegCon")
batch5_sub = subset_samples(batch5, Type == "NegCon")
batch6_sub = subset_samples(batch6, Type == "NegCon")
batch7_sub = subset_samples(batch7, Type == "NegCon")
batch8_sub = subset_samples(batch8, Type == "NegCon")
batch9_sub = subset_samples(batch9, Type == "NegCon")
batch10_sub = subset_samples(batch10, Type == "NegCon")

batch1_sub <- prune_taxa(taxa_sums(batch1_sub) > 0, batch1_sub)
batch2_sub <- prune_taxa(taxa_sums(batch2_sub) > 0, batch2_sub)
batch3_sub <- prune_taxa(taxa_sums(batch3_sub) > 0, batch3_sub)
batch4_sub <- prune_taxa(taxa_sums(batch4_sub) > 0, batch4_sub)
batch5_sub <- prune_taxa(taxa_sums(batch5_sub) > 0, batch5_sub)
batch6_sub <- prune_taxa(taxa_sums(batch6_sub) > 0, batch6_sub)
batch7_sub <- prune_taxa(taxa_sums(batch7_sub) > 0, batch7_sub)
batch8_sub <- prune_taxa(taxa_sums(batch8_sub) > 0, batch8_sub)
batch9_sub <- prune_taxa(taxa_sums(batch9_sub) > 0, batch9_sub)
batch10_sub <- prune_taxa(taxa_sums(batch10_sub) > 0, batch10_sub)


# Make the taxa names into a vector so you can remove them, then use the keep vector for the prune taxa argument, because it wants the argument to be true (matching), and repeat for all batches
batch1_ctrl <- as.vector(taxa_names(batch1_sub)) 
batch1_vec <- as.vector(taxa_names(batch1)) 
batch1_kp <- setdiff(batch1_vec, batch1_ctrl)
batch1_clean <- prune_taxa(batch1_kp, batch1)

batch2_ctrl <- as.vector(taxa_names(batch2_sub)) 
batch2_vec <- as.vector(taxa_names(batch2)) 
batch2_kp <- setdiff(batch2_vec, batch2_ctrl)
batch2_clean <- prune_taxa(batch2_kp, batch2)

batch3_ctrl <- as.vector(taxa_names(batch3_sub)) 
batch3_vec <- as.vector(taxa_names(batch3)) 
batch3_kp <- setdiff(batch3_vec, batch3_ctrl)
batch3_clean <- prune_taxa(batch3_kp, batch3)

batch4_ctrl <- as.vector(taxa_names(batch4_sub)) 
batch4_vec <- as.vector(taxa_names(batch4)) 
batch4_kp <- setdiff(batch4_vec, batch4_ctrl)
batch4_clean <- prune_taxa(batch4_kp, batch4)

batch5_ctrl <- as.vector(taxa_names(batch5_sub)) 
batch5_vec <- as.vector(taxa_names(batch5)) 
batch5_kp <- setdiff(batch5_vec, batch5_ctrl)
batch5_clean <- prune_taxa(batch5_kp, batch5)

batch6_ctrl <- as.vector(taxa_names(batch6_sub)) 
batch6_vec <- as.vector(taxa_names(batch6)) 
batch6_kp <- setdiff(batch6_vec, batch6_ctrl)
batch6_clean <- prune_taxa(batch6_kp, batch6)

batch7_ctrl <- as.vector(taxa_names(batch7_sub)) 
batch7_vec <- as.vector(taxa_names(batch7)) 
batch7_kp <- setdiff(batch7_vec, batch7_ctrl)
batch7_clean <- prune_taxa(batch7_kp, batch7)

batch8_ctrl <- as.vector(taxa_names(batch8_sub)) 
batch8_vec <- as.vector(taxa_names(batch8)) 
batch8_kp <- setdiff(batch8_vec, batch8_ctrl)
batch8_clean <- prune_taxa(batch8_kp, batch8)

batch9_ctrl <- as.vector(taxa_names(batch9_sub)) 
batch9_vec <- as.vector(taxa_names(batch9)) 
batch9_kp <- setdiff(batch9_vec, batch9_ctrl)
batch9_clean <- prune_taxa(batch9_kp, batch9)

batch10_ctrl <- as.vector(taxa_names(batch10_sub)) 
batch10_vec <- as.vector(taxa_names(batch10)) 
batch10_kp <- setdiff(batch10_vec, batch10_ctrl)
batch10_clean <- prune_taxa(batch10_kp, batch10)

# Combine the phyloseq objects of ID'd contaminants into one file
phy_trim <- merge_phyloseq(batch1_sub, batch2_sub, batch3_sub, batch4_sub, batch5_sub, batch6_sub, batch7_sub, batch8_sub, batch9_sub, batch10_sub) 

# Save the phyloseq object of identified contaminants in case further examination is desired
saveRDS(phy_trim, "Data/Outputs/2025-06-17/COIB/InterimObjects/DecontamRemoved.rds")
```

```{r trim_explore, echo = FALSE}
# Convert phyloseq of trimmed taxa to a dataframe
trimdf <- psmelt(phy_trim)

# Create a stacked bar plot of trimmed taxa by phylum by sample type per sample
trim_phyla <- ggplot(trimdf, aes(fill =Phylum, y = Abundance, x = Type)) +
  geom_bar(position = "stack", stat = "identity")

# Create a stacked bar plot of trimmed taxa by class by sample type per sample
trim_class <- ggplot(trimdf, aes(fill = Class, y = Abundance, x = Type)) +
  geom_bar(position = "stack", stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1))

# Save plots
ggsave("TrimmedPhyla.png", plot = trim_phyla, path = path_figB, width = 6, height = 4, units = "in")
ggsave("TrimmedClass.png", plot = trim_class, path = path_figB, width = 6, height = 4, units = "in")
```

```{r phy_prep, echo = FALSE}
# Merge the phyloseq objects of non-contaminants back together
# 
clean_phy <- merge_phyloseq(batch1_clean, batch2_clean, batch3_clean, batch4_clean, batch5_clean, batch6_clean, batch7_clean, batch8_clean, batch9_clean, batch10_clean) 

# List the phyla identified in the phyloseq object - we want to remove non-target taxa
get_taxa_unique(clean_phy, taxonomic.rank = "Phylum")
```

```{r remove_nontarget, echo = FALSE}
# Remove non-target phyla (any bacteria, fungi, or algae) that were ID'd in the chunk above
clean_phy <- ps_clean %>% 
  subset_taxa(Phylum != "Ascomycota") 
clean_phy <- clean_phy %>% 
  subset_taxa(Phylum != "Bacillariophyta")
clean_phy <- clean_phy %>% 
  subset_taxa(Phylum != "Basidiomycota")
clean_phy <- clean_phy %>% 
  subset_taxa(Phylum != "Chlorophyta")
clean_phy <- clean_phy %>% 
  subset_taxa(Phylum != "Oomycota")
clean_phy <- clean_phy %>% 
  subset_taxa(Phylum != "Pseudomonadota")
clean_phy <- clean_phy %>% 
  subset_taxa(Phylum != "Rhodophyta")
clean_phy <- clean_phy %>% 
  subset_taxa(Phylum != "Rotifera")
clean_phy <- clean_phy %>% 
  subset_taxa(Phylum != "Tubulinea")
clean_phy <- clean_phy %>% 
  subset_taxa(Phylum != "Streptophyta")
clean_phy <- clean_phy %>% 
  subset_taxa(Class != "Chrysophyceae")
clean_phy <- clean_phy %>% 
  subset_taxa(Class != "Raphidophyceae")

clean_phy <- clean_phy %>%
  subset_taxa(Genus != "Homo") # Remove human DNA

# Clean out taxa/SV columns that are no longer present
clean_phy <- prune_taxa(taxa_sums(clean_phy) > 0, clean_phy) 
clean_phy <- prune_samples(sample_sums(clean_phy) > 0, clean_phy)

# Save the cleaned phyloseq object
saveRDS(clean_phy, "Data/Outputs/2025-06-17/COIB/InterimObjects/CleanPhyloseq.RDS")
```

